---
title: "Exploring US Census data in R with tidycensus"
subtitle: "ISDS R Users Group Webinar"
date: "February 19, 2018"
author: "Kyle Walker, TCU"
output: 
  xaringan::moon_reader: 
    seal: yes
  
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
library(tidycensus)
library(tidyverse)
options(tigris_use_cache = TRUE)
```

## Follow along!  

* Slides: URL

* Code: Gist

---

## About me

* Work: professor at TCU/spatial data science consultant

* Research: urban geography, spatial demography, open data science

* Software: tidycensus, tigris, idbr

* Forthcoming book: _Analyzing the US Census with R_

---

## Disclaimer

This webinar uses Census Bureau data but is not endorsed or certified by the Census Bureau.  

---

## What we'll cover

* Census data: the basics

* Acquiring data with tidycensus

* Margins of error in tidycensus

* Mapping with tidycensus and ggplot2

---

## Census data: tables and features

Work with US Census data commonly includes two components: 

* __Data tables__ obtained from the decennial Census, American Community Survey, or other sources

* __Geographic features__ obtained from the Census's TIGER/Line database

---

## Census tables: American FactFinder

<img src=img/factfinder.PNG style="width: 700px">

---

## Census geography: TIGER/Line shapefiles

<img src=img/dropdown.PNG style="width: 400px">

---

## tidycensus: get Census data in R

* R package first released in mid-2017

* Allows R users to obtain decennial Census and ACS data pre-formatted for use with tidyverse tools (dplyr, ggplot2, etc.)

* Optionally returns geographic data as simple feature geometry for common Census geographies

---

## What tidycensus can do

```{r, fig.width=5}
tx <- get_acs(geography = "county", state = "TX", 
              variables = "B19013_001", geometry = TRUE)

ggplot(tx, aes(fill = estimate)) + geom_sf()
```

---

## tidycensus: the basics

* The `get_decennial()` and `get_acs()` functions give access to the decennial Census and American Community Survey, respectively

* Required arguments: `geography` and `variables`

* Default `year`: 2010 (decennial Census) and 2012-2016 (ACS)

---

## How tidycensus works

* tidycensus formats your arguments to make a request to the appropriate Census or ACS Application Programming Interface (API)

* A __tibble__ (or sf tibble) is returned in tidy (long) format by default containing the requested data

---

## Census geography

<img src=img/hierarchy.PNG style="width: 550px">

---

## Census variables

* __Variables__ in tidycensus are identified by their Census ID, e.g. `B19013_001`

* Entire __tables__ of variables can be requested with the `table` argument, e.g. `table = "B19001"`

* Users can request multiple variables at a time, and set custom names with a named vector

---

## Searching for Census variables

* Variable definitions for a given dataset can be loaded into R with the `load_variables()` function, and explored in RStudio with `View()`

```{r, eval = FALSE}
v16 <- load_variables(2016, "acs5", cache = TRUE)
View(v16)
```

---

## "Tidy" Census data

```{r}
income <- get_acs(geography = "state", table = "B19001")
income
```

---

## "Wide" Census data

```{r}
inc_wide <- get_acs(geography = "state", table = "B19001", 
                    output = "wide")
inc_wide
```

---

class: middle, center, inverse

## Margins of error in tidycensus

---

## Margins of error in the ACS

* American Community Survey: _sample_ of approximately 3 million American households per year

* Geographies of population > 65,000 available in the 1-year ACS; all geographies (starting with block groups) available in the 5-year ACS

* ACS __estimates__ associated with __margins of error__; default confidence level of 90 percent

---

## Margins of error in `get_acs()`


```{r}
az <- get_acs(geography = "county", variables = "B19013_001", 
              state = "AZ")

head(az)
```

---

## Visualizing margins of error

```{r, eval = FALSE}
az %>%
  mutate(NAME = str_replace(NAME, " County, Arizona", "")) %>%
  ggplot(aes(x = estimate, y = reorder(NAME, estimate))) + 
  geom_errorbarh(aes(xmin = estimate - moe, xmax = estimate + moe)) +
  geom_point(color = "red", size = 3) +
  scale_x_continuous(labels = scales::dollar) + 
  labs(title = "Household income by county in Arizona",
       subtitle = "2012-2016 American Community Survey",
       y = "",
       x = "ACS estimate (bars represent margin of error)")
```


---

## Visualizing margins of error

<img src=img/az.png style="width: 800px">

---

## Derived margins of error in tidycensus

* Margins of error for derived estimates available with the `moe_sum()`, `moe_prop()`, `moe_ratio()`, and `moe_product()` functions

* When possible, attempt to locate pre-computed derived estimates in the ACS before computing yourself (e.g. in Data Profile or Subject Tables)

---
class: middle, center, inverse

## Mapping data obtained with tidycensus

---

## Census "geometry" in R

* __tigris__ package: enables users to obtain and load Census geography as R objects

* __sf__ package: next-generation model for representing vector spatial data in R as _simple features_

Example: 

```{r}
library(tigris)
mi <- counties("MI", cb = TRUE, class = "sf")
```


---

## Census "geometry" in R

```{r}
plot(mi$geometry)
```

---

## Geometry in tidycensus

* For common geographies (`"state"`, `"county"`, `"tract"`, `"block group"`, `"block"`, and `"zcta"`) tidycensus can load simple feature geometry with the argument `geometry = TRUE`. 

Example: 

```{r}
cook <- get_acs(geography = "tract", state = "IL", county = "Cook", 
                variables = c(hhincome = "B19013_001"), geometry = TRUE)
```

---

## Geometry in tidycensus

```{r}
head(cook)
```


---

## Mapping Census data: __ggplot2__

* The `geom_sf()` function in ggplot2 (development version) allows for the mapping of simple features objects

Example: 

```{r, eval = FALSE}
library(viridis)
ggplot(cook, aes(fill = estimate, color = estimate)) + 
  geom_sf() + 
  theme_minimal() + 
  coord_sf(crs = 26916, datum = NA) + 
  scale_color_viridis(option = "cividis", guide = FALSE) + 
  scale_fill_viridis(option = "cividis", 
                     labels = scales::dollar) + 
  labs(title = "Median household income, 2012-2016 ACS", 
       subtitle = "Census tracts in Cook County, Illinois", 
       fill = "")
```

---

## Mapping Census data: __ggplot2__

<img src=img/cook.png style="width: 800px">

---

## Mapping Census data: __mapview__

```{r}
library(mapview)
mapview(cook, zcol = "estimate", legend = TRUE)
```

---

## Interactive exploration: __leaflet__

<iframe src="https://walkerke.github.io/urbanslides/chicago/img/il_income.html" height = "450" width = "800" frameborder="0" scrolling="no"></iframe>

.footnote[[Click here for source code](https://gist.github.com/walkerke/2d534dc0dd638ccdbaeef1ca83f4fe86)]
---

## Interactive exploration: __plotly__

<iframe src="brushing.html" height = "500" width = "1000" frameborder="0" scrolling="no"></iframe>

.footnote[[Click here for source code](https://gist.github.com/walkerke/93bfe80bb7735aa6265a61013eaed3fa)]

---

## Thank you!

For more: 

* Hire me as a consultant: kwalkerdata@gmail.com

* Take my DataCamp course on US Census data in R - coming this spring

* Join my mailing list: http://eepurl.com/cPGKZD 

* Follow me on Twitter: @kyle_e_walker









<style>

h1, h2, h3 {
  color: #386890; 
}

a {
  color: #90b4d2; 
}

.inverse {
  background-color: #386890; 

}
</style>







